---
title: "Clustering"
author: "Group 10 (Umar, Cory, Caroline, Benji)"
date: "10/9/2022"
output:
  pdf_document: default
  html_notebook: default
---

Source: https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset

This is a dataset based off of 70,000 records of patient data (Heart Related).
Columns (13): ID, Age, Height(cm), Weight(kg), Gender, Systolic Blood Pressure 
(AP_HIGH), Diastolic Blood Pressure (AP LOW), Cholesterol, Glucose, Smoking, 
Alcohol Intake, Physical Activity, Presence or Absence of cardiovascular disease.

The .csv file needed to be edited a bit in Microsoft Excel before using it in R.
I just performed a split column delimiter function around semicolons, to divide
the singular column that existed into 13. Each row had 13 variables in 1 column 
separated by semicolons, the function I ran split it up into 13 columns, making
a 70,000 x 13 table.

https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings

Visit the website above to better understand Systolic and Diastolic Blood Pressure


### Cleaning Data


```{r}


library(ggplot2)
library(flexclust)
library(mclust)


# Read in .csv file
heart <- read.csv("cardio_train.csv")


# Clean out any rows that have an unrealistic blood pressure (AP_HIGH & AP_LOW)
# They looked to be input errors by the person who made the data set
s <- subset(heart, AP_HIGH > 50)

s1 <- subset(s, AP_HIGH < 200)

s2 <- subset(s1, AP_LOW > 25)

s3 <- subset(s2, AP_LOW < 200)


# Removing little people(4'10") and Giants(7'3"), values are in cm
s4 <- subset(s3, HEIGHT > 147)

s5 <- subset(s4, HEIGHT < 220)


# Removing anyone below 90 lbs and above 375 lbs, the values are in kg
s6 <- subset(s5, WEIGHT > 40)

h1 <- subset(s6, WEIGHT < 180)

# AGE is in days so to get years i just divide by 365
h1$AGE <- (h1$AGE / 365)

# Removing people under 40
h <- subset(h1, AGE > 39)

# Checking for any NA values
# There is none
colSums(is.na(h))

# Everything that should be factored is factored
h$GENDER <- factor(h$GENDER)
h$CHOLESTEROL <- factor(h$CHOLESTEROL)
h$GLUCOSE <- factor(h$GLUCOSE)
h$SMOKE <- factor(h$SMOKE)
h$ALCOHOL <- factor(h$ALCOHOL)
h$PHYSICAL_ACTIVITY <- factor(h$PHYSICAL_ACTIVITY)
h$CARDIO_DISEASE <- factor(h$CARDIO_DISEASE)


# There is now 67,685 rows

str(h)

```

### Kmean Clusters

Determining how many numbers of clusters to use for the data

```{r}

# Function for finding optimal number of clusters

hplot <- function(data, nc = 15, seed = 2354)
{
  hss <- (nrow(h) - 1) * sum(apply(h, 2, var))
  for (i in 2:nc)
  {
    set.seed(seed)
    hss[i] <- sum(kmeans(h, centers = i)$withinss)
  }
  plot(1:nc, hss, type = "b", xlab = "Number of Clusters",
       ylab = "Within groups sum of squares")
}

# Show graph

hplot(h)

cat("Looking at the graph, three seems to be the optimal number of clusters.\n
That's where the 'elbow' is.")

```

Making the kmean clusters based off of 3 clusters

```{r}

# Set unique and repeatable random variable

set.seed(2354)

# Clustering off of Systolic and Diastolic blood pressure

heartCluster <- kmeans(h[, 6:7], 3, nstart = 20)

# Data is too large to show the summary for

cat("K-means clustering with 3 clusters of sizes 12542, 18119, 37024

Cluster means:
   AP_HIGH   AP_LOW
1 106.5830 70.24629
2 148.3140 90.72096
3 122.4537 80.46778

Clustering vector:

[ reached getOption('max.print') -- omitted 66685 entries ]

Within cluster sum of squares by cluster:
[1] 1330107 3536136 1932918
 (between_SS / total_SS =  71.9 %)

Available components:

[1] 'cluster'      'centers'      'totss'        'withinss'    
[5] 'tot.withinss' 'betweenss'    'size'         'iter'        
[9] 'ifault'   ")

```

Looking at clusters through the lens of having heart disease

```{r}

# Comparing the cluster and the presence of heart disease
# A '1' on the x-axis means they have heart disease

table(heartCluster$cluster, h$CARDIO_DISEASE)

```

Displaying clusters on scatter plot

```{r}

# Plotting the clusters on a scatter plot

plot(h$AP_HIGH, h$AP_LOW, pch = 21, bg = c("red", "pink", "slateblue")
     [unclass(heartCluster$cluster)], main = "Heart Data")

```

### Hierarchical Clustering

```{r}

# Removing any categorical data and ID's

hrt <- subset(h, select = -c(ID, GENDER, CHOLESTEROL, GLUCOSE, SMOKE, ALCOHOL,
                             PHYSICAL_ACTIVITY, CARDIO_DISEASE))

# Limiting data points to 50 for the sake of the visibility of the table

hrt <- hrt[1:50,]

# Displaying new table

head(hrt)

# Scaling the data

hrt.scaled <- scale(hrt)

head(hrt.scaled)

```

Displaying Hierarchical Graph (Dendogram)

```{r}

# Finding the distances between each data point

di <- dist(hrt.scaled)

fit.average <- hclust(di, method = "average")

plot(fit.average, hang = -1, cex = .8, main = "Hierarchical Clustering")

```

Cutting data

```{r}

# Learning more about the data through cutting

# SIDENOTE: Hierarchical Clustering isn't the easiest to understand, but is good 
# for learning more about your data

for (c in 3:11)
{
  cluster_cut <- cutree(fit.average, c)
  table_cut <- table(cluster_cut, hrt$AP_LOW)
  print(table_cut)
  ri <- randIndex(table_cut)
  print(paste("cut=", c, "Rand Index = ", ri))
}

```

### Model Based

```{r}

# Shrink data set to 500 for the sake of the algorithms

hhrt <- subset(h[1:500,], select = -c(ID, GENDER, CHOLESTEROL, GLUCOSE, SMOKE, ALCOHOL,
                             PHYSICAL_ACTIVITY, CARDIO_DISEASE))

fitt <- Mclust(hhrt)

# Plot results

plot(fitt)

# Display the best model

summary(fitt)

cat("The first one is the BIC graph, it suggests EEV with 7 groups, based on the 
\nhighest BIC value and the number of components it intersects with. The summary
\nof the model also suggests the same thing. Classification, uncertainty, and
\ndensity all have a very similar looking graphs. They just show correlation
\non a scatterplot matrix. Unsurprisingly, age and height seem to have very
\nlittle correlation, due to the age range of the data (40 - 65), and systolic
\nand diastolic blood pressure have the most correlation.")

```